---
title: "Join spatial data with a data frame"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set()
```


## Set up

The packages we'll use for this adventure are:

```{r import_packages, exercise = TRUE, exercise.lines = 10, exercise.eval = TRUE}

library(sf) # working with geographic data

library(tidyverse) # plotting and data wrangling

library(mapview) # plotting and exploring geographic data

library(gapminder) # gapminder data
```


## Reading in the data


```{r import_packages2, echo = FALSE, message = FALSE, warning = FALSE}

library(sf) # working with geographic data

library(tidyverse) # plotting and datawrangling

library(mapview) # plotting and exploring geographic data

library(gapminder) # gapminder data
```


### Data frame

The `gapminder` data is read in when the `gapminder` package is loaded. We can prepare the data by filtering out just the data for the continent "Africa".

```{r read_data, exercise = TRUE, exercise.eval = TRUE}

africa_gap <- gapminder %>%
                  filter(continent == "Africa")

```

We can inspect the data using `head()`, `summary()`, and `names()`

```{r head_data, exercise = TRUE}

head(africa_gap)

```

```{r summary_data, exercise = TRUE}

summary(africa_gap)

```

```{r names_data, exercise = TRUE}

names(africa_gap)

```


```{r country_column_quiz, echo = FALSE}
quiz(
  question("Which column contains the country name?",
    answer("continent"),
    answer("year"),
    answer("country", correct = TRUE),
    answer("gdpPercap"),
    incorrect = "Hint: Look at the head of the data frame or use names(africa_gap) to find out the column names.",
    allow_retry = TRUE
  )
)
```

### Spatial data

Read in the continent data `africa_map.shp` which contains a subset of data from the `rnaturalearth` package for the continent, Africa. 

```{r read_spatial_data, exercise = TRUE, exercise.eval = TRUE}

africa_map <- st_read("data/africa_map.shp")

```


Doublecheck that the class is `sf` using the function `class`. 

```{r check_class, exercise = TRUE, exercise.eval = TRUE}

class(africa_map)
```

Check the head of the data


```{r head_africa_map, exercise=TRUE, exercise.eval = TRUE}

head(africa_map)

```


### Exercise 1

```{r quiz, echo = FALSE}
quiz(
  question("In which column do you find the spatial information?",
    answer("country"),
    answer("iso_a3"),
    answer("geometry", correct = TRUE),
    answer("continent"),
    incorrect = "Hint: Look at the head of the spatial data frame using head(africa_map), which column contains coordinates?",
    allow_retry = TRUE
  )
)
```


### Exercise 2


```{r quiz2}

quiz(
  question("What type of geometry is the spatial file?",
    answer("point"),
    answer("line"),
    answer("polygon"),
    answer("multipolygon", correct = TRUE),
    incorrect = "Hint: Look at geometry column of the dataset, what type of spatial data does it report?",
    allow_retry = TRUE
  )
)
```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Joining the datasets

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz3}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
